---
title: "SML Assigment"
author: "Joachim SINYABE DANBE"
date: "2025-12-06"
output:
  pdf_document:
    latex_engine: pdflatex
    keep_tex: true
---





```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(data.table)
library(dplyr)
``` 

#Llet's load CMAPSS data
```{r}
#Dataset : NASA C-MAPSS (Commercial Modular Aero-Propulsion System Simulation)
#Source : \url{https://www.kaggle.com/datasets/palbha/cmapss-jet-engine-simulated-data

col_names <- c("engine_id", "cycle", "op_setting_1", "op_setting_2", "op_setting_3", paste0("sensor_", 1:21))

load_cmapss <- function(file_path) {
  read.table(file_path, col.names = col_names
  ) %>%arrange(engine_id, cycle)
}


path ="/home/guest/Desktop/JSD/SML/Dataset/"

train_raw <- load_cmapss(paste0(path, "train_FD001.txt"))
test_raw  <- load_cmapss(paste0(path, "test_FD001.txt"))
rul_actual   <- read.table(paste0(path, "RUL_FD001.txt"), header = FALSE)

colnames(rul_actual) <- "RUL"

colnames(test_raw)  <- col_names
colnames(rul_actual)   <- "RUL"
```



#Wheel 1 # Formulation problem

#Wheel 2 # Data explonatory

# We add RUL to training data and for test data we add just the last observation
```{r}
head(train_raw)
head(test_raw)
head(rul_actual)
train_raw <- train_raw %>%
    group_by(engine_id) %>%
    mutate(
      max_cycle = max(cycle),
      rul = max_cycle - cycle
    ) %>%
    ungroup() %>%
    select(-max_cycle)
test_raw <- test_raw %>%
    group_by(engine_id) %>%
    filter(cycle == max(cycle)) %>%
    ungroup() %>%
    bind_cols(rul_actual)
```

#The informations about the data set are  

```{r}
cat("Train:", nrow(train_raw), "observations,",
n_distinct(train_raw$engine_id), "engines\n")
cat("Test:", nrow(test_raw), "observations,",
n_distinct(test_raw$engine_id), "engines\n")
cat("Mean RUL:", round(mean(train_raw$rul), 1), "cycles\n\n")

```

#RUL Distribution

```{r}
library(ggplot2)
#install.packages("gridExtra")
library(gridExtra)

p1 <- ggplot(train_raw, aes(x = rul)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  labs(title = "Distribution of RUL (Training Set)",
       x = "RUL (cycles)", 
       y = "Frequency")

p2 <- train_raw %>%
  group_by(engine_id) %>%
  summarise(max_rul = max(rul), .groups = "drop") %>%
  ggplot(aes(x = max_rul)) +
  geom_histogram(bins = 30, fill = "coral", alpha = 0.7) +
  labs(title = "Distribution of Engine Lifetimes",
       x = "Total Lifetime (cycles)", 
       y = "Count")

grid.arrange(p1, p2, ncol = 2)

```
# Descriptives statistics of key sensors

```{r}
train_raw %>%
  select(rul, sensor_2, sensor_4, sensor_11, sensor_7, sensor_21) %>%
  summary() %>%
  knitr::kable(caption = "Descriptive Statistics of Key Variables")

```

# constant sensor and Correlation matrix

```{r}
library(dplyr)
library(tidyr)

sensor_variance <- train_raw %>%select(starts_with("sensor_")) %>%summarise(across(everything(),
      ~ var(.x, na.rm = TRUE)
    )
  ) %>%pivot_longer(everything(),
    names_to = "sensor",
    values_to = "variance"
  ) %>%arrange(variance)

# constant with sensors (variance < 0.001)
constant_sensors <- sensor_variance %>%
  dplyr::filter(variance < 0.001) %>%
  dplyr::pull(sensor)

cat("Constant sensors to remove:", paste(constant_sensors, collapse = ", "), "\n")
cat("Number of sensors to remove:", length(constant_sensors), "\n")

# variance table
sensor_variance %>%
  dplyr::mutate(variance = round(variance, 4)) %>%
  head(10) %>%
  knitr::kable(caption = "Sensor Variance (Top 10 Lowest)")

```




```{r}
train_raw %>%
  filter(engine_id %in% 1:5) %>%
  ggplot(aes(x = cycle, y = rul, color = as.factor(engine_id))) +
  geom_line(size = 1, alpha = 0.8) +
  labs(title = "RUL Trajectories for 5 Engines",
       x = "Cycle", 
       y = "RUL (cycles remaining)",
       color = "Engine ID") +
  theme_minimal()
```


#Wheel 3: Feature Engineering & Model Specification


```{r}
add_features <- function(df) {
  df %>%
    dplyr::group_by(engine_id) %>%
    dplyr::arrange(cycle) %>%
    dplyr::mutate(
      sensor_2_roll_mean = zoo::rollmean(sensor_2, k = 5, fill = NA, align = "right"),
      sensor_4_roll_mean = zoo::rollmean(sensor_4, k = 5, fill = NA, align = "right"),
      sensor_11_roll_mean = zoo::rollmean(sensor_11, k = 5, fill = NA, align = "right")
    ) %>%
    dplyr::ungroup()
}


train_fe <- add_features(train_raw) %>%
  select(-all_of(constant_sensors))  # Remove constant sensors

cat("Features after engineering:", ncol(train_fe) - 3, "\n")  # -3 for id, cycle, rul
cat("Observations after removing NA:", nrow(train_fe), "\n")

colnames(train_fe)

```

# Visualization of Feature Engineering Impact

```{r}
train_fe %>%
  dplyr::filter(engine_id == 1) %>%
  dplyr::select(cycle, sensor_2, sensor_2_roll_mean) %>%
  tidyr::pivot_longer(-cycle, names_to = "feature", values_to = "value") %>%
  ggplot2::ggplot(aes(x = cycle, y = value, color = feature)) +
  ggplot2::geom_line(size = 0.8, alpha = 0.8) +
  ggplot2::labs(
    title = "Feature Engineering: Sensor 2 (Engine #1)",
    x = "Cycle", y = "Value", color = "Feature"
  ) +
  ggplot2::theme_minimal()

```


#Wheel 4 Loss function 

**Mean Squared Error (MSE)**:
$$L(y, \hat{y}) = (y - \hat{y})^2$$


**Empirical Risk**:
$$R_{\text{emp}}(\theta) = \frac{1}{n}\sum_{i=1}^{n}(y_i - f_\theta(\mathbf{x}_i))^2$$

**Evaluation Metric**: RMSE = $\sqrt{\text{MSE}}$

```{r}
rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2))
}

mae <- function(actual, predicted) {
  mean(abs(actual - predicted))
}

r2_score <- function(actual, predicted) {
  1 - sum((actual - predicted)^2) / sum((actual - mean(actual))^2)
}
```



#Wheel 5 : Learning  criteria & Optimisation
##Train/VAlidation split

```{r}
unique_engines <- unique(train_fe$engine_id)
n_train_engines <- floor(0.8 * length(unique_engines))

train_engine_ids <- sample(unique_engines, n_train_engines)
val_engine_ids <- setdiff(unique_engines, train_engine_ids)


train_data <- train_fe %>% filter(engine_id %in% train_engine_ids)
val_data <- train_fe %>% filter(engine_id %in% val_engine_ids)


X_train <- train_data %>% 
  select(-engine_id, -cycle, -rul) %>%
  as.data.frame()
y_train <- train_data$rul

X_val <- val_data %>% 
  select(-engine_id, -cycle, -rul) %>%
  as.data.frame()
y_val <- val_data$rul

cat("Training set:", nrow(X_train), "observations\n")
cat("Validation set:", nrow(X_val), "observations\n")
cat("Number of features:", ncol(X_train), "\n")
```
```{r}

na_count_train <- sum(is.na(X_train))
na_count_val <- sum(is.na(X_val))

if (na_count_train > 0) {
  cat("Warning: NA values in training data:", na_count_train, "\n")
  X_train <- na.omit(X_train)
  y_train <- y_train[complete.cases(train_data %>% select(-engine_id, -cycle, -rul))]
}

if (na_count_val > 0) {
  cat("Warning: NA values in validation data:", na_count_val, "\n")
  X_val <- na.omit(X_val)
  y_val <- y_val[complete.cases(val_data %>% select(-engine_id, -cycle, -rul))]
}

# Check for infinite values
if (any(!is.finite(as.matrix(X_train)))) {
  cat("Warning: Infinite values detected in training data\n")
  X_train <- X_train %>% mutate(across(everything(), ~replace(., !is.finite(.), NA))) %>% na.omit()
}

cat("\nFinal dimensions:\n")
cat("  X_train:", nrow(X_train), "×", ncol(X_train), "\n")
cat("  X_val:", nrow(X_val), "×", ncol(X_val), "\n\n")
```

### Model 1:  Linear  Regression

```{r}
library(caret)

cat("Training Linear Regression...\n")
model_lm <- lm(y_train ~ ., data = X_train)

#predictions 
pred_lm_train <- predict(model_lm, X_train)
pred_lm_val <- predict(model_lm, X_val)


# Metrics
rmse_lm_train <- rmse(y_train, pred_lm_train)
rmse_lm_val <- rmse(y_val, pred_lm_val)
r2_lm_val <- r2_score(y_val, pred_lm_val)


cat("Linear Regression Results:\n")
cat("  Features used:", ncol(X_train), "\n")
cat("  RMSE (train):", round(rmse_lm_train, 2), "\n")
cat("  RMSE (val):", round(rmse_lm_val, 2), "\n")
cat("  R² (val):", round(r2_lm_val, 3), "\n\n")

```
# wheel 6 : Regularisation 
## Model 2: Ridge Regression

```{r}
#install.packages("glmnet")
library(glmnet)

cat("Training Ridge Regression with CV...\n")

X_train_matrix <- as.matrix(X_train)
X_val_matrix <- as.matrix(X_val)

set.seed(123)
cv_ridge <- cv.glmnet(X_train_matrix, y_train, alpha = 0, nfolds = 5)

lambda_opt <- cv_ridge$lambda.min

# Train with optimal lambda
model_ridge <- glmnet(X_train_matrix, y_train, alpha = 0, lambda = lambda_opt)

# Predictions
pred_ridge_train <- predict(model_ridge, X_train_matrix)[,1]
pred_ridge_val <- predict(model_ridge, X_val_matrix)[,1]

# Metrics
rmse_ridge_train <- rmse(y_train, pred_ridge_train)
rmse_ridge_val <- rmse(y_val, pred_ridge_val)
r2_ridge_val <- r2_score(y_val, pred_ridge_val)

cat("Ridge Regression Results:\n")
cat("  Optimal \\lamdda:", round(lambda_opt, 4), "\n")
cat("  RMSE (train):", round(rmse_ridge_train, 2), "\n")
cat("  RMSE (val):", round(rmse_ridge_val, 2), "\n")
cat("  R² (val):", round(r2_ridge_val, 3), "\n\n")

```



#Visualisation : Lambda Selection (Ridge)
```{r}
plot(cv_ridge, main = "")
abline(v = log(lambda_opt), col = "red", lty = 2, lwd = 2)
```
## Model 3: Random Forest
```{r}
library(randomForest)

cat("Training Random Forest...\n")

set.seed(456)
model_rf <- randomForest(x = X_train, y = y_train, ntree = 500, mtry = floor(sqrt(ncol(X_train))), maxnodes = 50,importance = TRUE, nodesize = 5 )

# Predictions
pred_rf_train <- predict(model_rf, X_train)

pred_rf_val <- predict(model_rf, X_val)

# Metrics
rmse_rf_train <- rmse(y_train, pred_rf_train)
rmse_rf_val <- rmse(y_val, pred_rf_val)
r2_rf_val <- r2_score(y_val, pred_rf_val)

cat("Random Forest Results:\n")
cat("  RMSE (train):", round(rmse_rf_train, 2), "\n")
cat("  RMSE (val):", round(rmse_rf_val, 2), "\n")
cat("  R² (val):", round(r2_rf_val, 3), "\n\n")

```
# top 15 features importants in random Forest
```{r}
importance_df <- importance(model_rf) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("feature") %>%
  dplyr::arrange(dplyr::desc(`%IncMSE`)) %>%
  head(15)

ggplot(importance_df, aes(x = reorder(feature, `%IncMSE`), y = `%IncMSE`)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  coord_flip() +
  labs(title = "Top 15 Feature Importance (Random Forest)",
       x = "Feature",
       y = "% Increase in MSE when permuted") +
  theme_minimal()
```



# Wheel 7: Extrinsic Predictive Comparisons
## Comparison Table
```{r model_xgboost}

results <- tibble(
  Model = c("Baseline (Mean)", "Linear Regression", "Ridge", 
            "Random Forest"),
  RMSE_Train = c(
    rmse(y_train, rep(mean(y_train), length(y_train))),
    rmse_lm_train,
    rmse_ridge_train,
    rmse_rf_train
  ),
  RMSE_Val = c(
    rmse(y_val, rep(mean(y_train), length(y_val))),
    rmse_lm_val,
    rmse_ridge_val,
    rmse_rf_val
  ),
  R2_Val = c(
    0,
    r2_lm_val,
    r2_ridge_val,
    r2_rf_val
  ),
  MAE_Val = c(
    mae(y_val, rep(mean(y_train), length(y_val))),
    mae(y_val, pred_lm_val),
    mae(y_val, pred_ridge_val),
    mae(y_val, pred_rf_val)
  )
) %>%
  mutate(
    RMSE_Train = round(RMSE_Train, 2),
    RMSE_Val = round(RMSE_Val, 2),
    R2_Val = round(R2_Val, 3)
  )

knitr::kable(results, 
             caption = "Model Comparison on Validation Set",
             align = "lcccc")


best_model_idx <- which.min(results$RMSE_Val[-1]) + 1  # Exclude baseline
best_model_name <- results$Model[best_model_idx]
cat("\nBest performing model:", best_model_name, 
    "(RMSE =", results$RMSE_Val[best_model_idx], ")\n")

```
# visualisation of Model Performance Comparison
```{r}
results %>%
  select(Model, RMSE_Val, MAE_Val) %>%
  pivot_longer(-Model, names_to = "Metric", values_to = "Value") %>%
  ggplot(aes(x = reorder(Model, -Value), y = Value, fill = Metric)) +
  geom_col(position = "dodge", alpha = 0.8) +
  coord_flip() +
  labs(title = "Model Performance Comparison",
       x = "Model",
       y = "Error",
       fill = "Metric") +
  theme_minimal()
```

### Predictions vs Actual (RandomForest)

```{r}
#  When we se Random Forest predictions (typically best among the 3)

pred_df <- tibble(
  actual = y_val,
  predicted = pred_rf_val,
  error = actual - predicted
)

p1 <- ggplot(pred_df, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", size = 1) +
  labs(title = "Predictions vs Actual (RandomForest)",
       x = "Actual RUL",
       y = "Predicted RUL") +
  theme_minimal()

p2 <- ggplot(pred_df, aes(x = error)) +
  geom_histogram(bins = 40, fill = "coral", alpha = 0.7) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", size = 1) +
  labs(title = "Prediction Error Distribution",
       x = "Error (Actual - Predicted)",
       y = "Frequency") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

## Wheel 8: Statistical Inference & Theoretical Justification

## Bootstrap Confidence Intervals

```{r}
cat("Computing Bootstrap Confidence Intervals (B=1000)...\n")

set.seed(999)
B <- 1000
bootstrap_rmse <- numeric(B)
bootstrap_r2 <- numeric(B)

for (b in 1:B) {
  idx <- sample(length(y_val), replace = TRUE)
  bootstrap_rmse[b] <- rmse(y_val[idx], pred_rf_val[idx])
  bootstrap_r2[b] <- r2_score(y_val[idx], pred_rf_val[idx])
}

# Calculate confidence intervals
ci_rmse_lower <- quantile(bootstrap_rmse, 0.025)
ci_rmse_upper <- quantile(bootstrap_rmse, 0.975)
ci_r2_lower <- quantile(bootstrap_r2, 0.025)
ci_r2_upper <- quantile(bootstrap_r2, 0.975)

cat("\nBootstrap 95% Confidence Intervals (RandomForest):\n")
cat("  RMSE:", round(rmse_rf_val, 2), 
    "  [", round(ci_rmse_lower, 2), ",", round(ci_rmse_upper, 2), "]\n")
cat("  R²:", round(r2_rf_val, 3), 
    "  [", round(ci_r2_lower, 3), ",", round(ci_r2_upper, 3), "]\n\n")

```


#Bootstrap Distribution
```{r vis11_bootstrap, fig.height=4}
boot_df <- tibble(
  RMSE = bootstrap_rmse,
  R2 = bootstrap_r2
)

p1 <- ggplot(boot_df, aes(x = RMSE)) +
  geom_histogram(bins = 40, fill = "skyblue", alpha = 0.7) +
  geom_vline(xintercept = rmse_rf_val, color = "red", size = 1) +
  geom_vline(xintercept = ci_rmse_lower, color = "blue", linetype = "dashed") +
  geom_vline(xintercept = ci_rmse_upper, color = "blue", linetype = "dashed") +
  labs(title = "Bootstrap RMSE Distribution",
       x = "RMSE",
       y = "Frequency",
       caption = "Red: observed | Blue: 95% CI") +
  theme_minimal()

p2 <- ggplot(boot_df, aes(x = R2)) +
  geom_histogram(bins = 40, fill = "lightgreen", alpha = 0.7) +
  geom_vline(xintercept = r2_rf_val, color = "red", size = 1) +
  geom_vline(xintercept = ci_r2_lower, color = "blue", linetype = "dashed") +
  geom_vline(xintercept = ci_r2_upper, color = "blue", linetype = "dashed") +
  labs(title = "Bootstrap R² Distribution",
       x = "R²",
       y = "Frequency") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```


## Statistical Significance Test

```{r wilcoxon_test}
errors_lm <- abs(y_val - pred_lm_val)
errors_rf <- abs(y_val - pred_rf_val)

test_result <- wilcox.test(errors_lm, errors_rf, paired = TRUE)

cat("Wilcoxon Signed-Rank Test (XGBoost vs Random Forest):\n")
cat("  Null hypothesis: No difference in prediction errors\n")
cat("  p-value:", format.pval(test_result$p.value, digits = 3), "\n")
cat("  Conclusion:", 
    ifelse(test_result$p.value < 0.05, 
           "RandomForests significantly better (p < 0.05)",
           "No significant difference (p > =  0.05)"), 
    "\n\n")
```
## Theoretical Justification
```{r}
d_vc <- 500  # Approximate for ensemble of trees
n <- nrow(X_train)
delta <- 0.05

bound_term <- sqrt((d_vc * log(2*n/d_vc) + log(4/delta)) / n)
r_true_upper <- rmse_rf_val + bound_term

cat("Generalization Bound (with 95% probability):\n")
cat("  Empirical RMSE:", round(rmse_rf_val, 2), "\n")
cat("  Bound term:", round(bound_term, 2), "\n")
cat("  True RMSE <= ", round(r_true_upper, 2), "\n\n")
```

# Wheel 9: Deployment & Practical Scalability

```{r}

example_engine_id <- val_engine_ids[1]

example_indices <- which(val_data$engine_id == example_engine_id)

pred_example <- pred_rf_val[example_indices]
actual_example <- y_val[example_indices]
cycles_example <- val_data$cycle[example_indices]

std_error <- sd(y_val - pred_rf_val)
ci_width <- 1.96 * std_error

pred_df_example <- tibble(
  cycle = cycles_example,
  rul_actual = actual_example,
  rul_predicted = pred_example,
  ci_lower = pred_example - ci_width,
  ci_upper = pred_example + ci_width
) %>%arrange(cycle)

cat("Deployment Example (Engine", example_engine_id, "):\n")
cat("  Total observations:", nrow(pred_df_example), "\n")
cat("  Cycle range:", min(cycles_example), "-", max(cycles_example), "\n")
cat("  Final predicted RUL:", round(tail(pred_example, 1), 1), "\n")
cat("  Final actual RUL:", round(tail(actual_example, 1), 1), "\n")
cat("  Absolute error:", round(abs(tail(pred_example, 1) - tail(actual_example, 1)), 1), "cycles\n\n")

```

### Visualization : Deployment Example with CI

```{r vis11_deployment, fig.height=5}
ggplot(pred_df_example, aes(x = cycle)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), 
              fill = "lightblue", alpha = 0.4) +
  geom_line(aes(y = rul_actual, color = "Actual"), size = 1.2) +
  geom_line(aes(y = rul_predicted, color = "Predicted"), size = 1.2) +
  labs(title = paste("JSD: Engine", example_engine_id),
       subtitle = "RUL Prediction with 95% Confidence Interval",
       x = "Cycle",
       y = "RUL (cycles)",
       color = "Type") +
  theme_minimal() +
  scale_color_manual(values = c("Actual" = "coral", "Predicted" = "steelblue"))

```

## Data Drift Monitoring

```{r drift_monitoring}
train_sensor_mean <- mean(train_data$sensor_2, na.rm = TRUE)
val_sensor_mean <- mean(val_data$sensor_2, na.rm = TRUE)

drift_pct <- abs(val_sensor_mean - train_sensor_mean) / train_sensor_mean * 100

cat("Data Drift Monitoring:\n")
cat("  Sensor 2 mean (train):", round(train_sensor_mean, 2), "\n")
cat("  Sensor 2 mean (val):", round(val_sensor_mean, 2), "\n")
cat("  Drift percentage:", round(drift_pct, 2), "%\n")
cat("  Alert threshold: 5%\n")
cat("  Status:", ifelse(drift_pct > 5, " ALERT - Retraining needed", 
                        "OK - No significant drift"), "\n\n")

```






#Exercise 2 : Practical Machine Learning - Digit Recogni-tion

```{r}
library(dslabs)
library(class)    
library(caret)     
library(pROC)     
library(ggplot2)
```


#Data exploratoration


```{r}
mnist <- read_mnist()

set.seed(19671210)

# we keep only digits 1 and 7
idx_train <- which(mnist$train$labels %in% c(1, 7))
idx_test  <- which(mnist$test$labels  %in% c(1, 7))

xtrain <- mnist$train$images[idx_train, ]
ytrain <- factor(mnist$train$labels[idx_train], levels = c(1, 7))
xtest  <- mnist$test$images[idx_test, ]
ytest  <- factor(mnist$test$labels[idx_test], levels = c(1, 7))


n_train <- nrow(xtrain)
n_test <- nrow(xtest)
p_feat  <- ncol(xtrain)

cat("Usefull informations\n")
cat("Training sample size (n):", n_train, "\n")
cat("Test sample size:", n_test, "\n")
cat("Feature dimensionality (p):", p_feat, "\n")
#cat("Ratio n/p:", round(n_train/p_feat, 3), "\n\n")

```
#Train k‑NN models (k = 1 and k = 27)


```{r}
library(ggplot2)
library(caret)

# 1NN confusion matrix plot
cm1 <- as.data.frame(conf_1nn$table)
colnames(cm1) <- c("True", "Predicted", "Freq")

ggplot(cm1, aes(x = Predicted, y = True, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 5) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Confusion Matrix - 1NN (k = 1)",
       x = "Predicted class",
       y = "True class") +
  theme_minimal()

# 27NN confusion matrix plot
cm27 <- as.data.frame(conf_27nn$table)
colnames(cm27) <- c("True", "Predicted", "Freq")

ggplot(cm27, aes(x = Predicted, y = True, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 5) +
  scale_fill_gradient(low = "white", high = "darkred") +
  labs(title = "Confusion Matrix - 27NN (k = 27)",
       x = "Predicted class",
       y = "True class") +
  theme_minimal()


cat("=== PERFORMANCE EVALUATION ===\n\n")

# 1NN Performance
cat("1NN (k = 1) Results\n")
conf_1nn <- confusionMatrix(pred_1nn, ytest, positive = "1")
print(conf_1nn$table)
cat("\nAccuracy:", round(conf_1nn$overall["Accuracy"], 4), "\n")
cat("Sensitivity (True Positive Rate):", round(conf_1nn$byClass["Sensitivity"], 4), "\n")
cat("Specificity (True Negative Rate):", round(conf_1nn$byClass["Specificity"], 4), "\n\n")

# 27NN Performance
cat("27NN (k = 27) Results\n")
conf_27nn <- confusionMatrix(pred_27nn, ytest, positive = "1")
print(conf_27nn$table)
cat("\nAccuracy:", round(conf_27nn$overall["Accuracy"], 4), "\n")
cat("Sensitivity (True Positive Rate):", round(conf_27nn$byClass["Sensitivity"], 4), "\n")
cat("Specificity (True Negative Rate):", round(conf_27nn$byClass["Specificity"], 4), "\n\n")


```

#ROC curves and AUC

```{r}
set.seed(19671210)

# train k-NN models via caret (gives class probabilities)
knn1  <- knn3(x = xtrain, y = ytrain, k = 1)
knn27 <- knn3(x = xtrain, y = ytrain, k = 27)

# predicted probabilities for class "1"
prob_1nn  <- predict(knn1,  xtest)[, "1"]
prob_27nn <- predict(knn27, xtest)[, "1"] 
roc_1nn  <- roc(response = ytest, predictor = prob_1nn,  levels = c("7", "1"))
roc_27nn <- roc(response = ytest, predictor = prob_27nn, levels = c("7", "1"))

auc_1nn  <- auc(roc_1nn)
auc_27nn <- auc(roc_27nn)
auc_1nn; auc_27nn

plot(roc_1nn,  col = "blue",  main = "ROC curves: 1-NN vs 27-NN")
plot(roc_27nn, col = "red", add = TRUE)
legend("bottomright",
       legend = c(paste0("1-NN  (AUC = ",  round(auc_1nn,  3), ")"),
                  paste0("27-NN (AUC = ", round(auc_27nn, 3), ")")),
       col = c("blue", "red"), lwd = 2)

```

#Error visualization: one FP and one FN


```{r}
# confusion for 27-NN to identify errors
conf_1nn$table
conf_27nn$table
# logical vectors
fp_idx <- which(pred_27nn == "1" & ytest == "7")
fn_idx <- which(pred_27nn == "7" & ytest == "1")

i_fp <- fp_idx[1]
i_fn <- fn_idx[1]


plot_digit <- function(x_vec, title = "") {
  m <- matrix(as.numeric(x_vec), nrow = 28, byrow = TRUE)
  image(t(apply(m, 2, rev)), col = gray(seq(0, 1, length = 256)),
        axes = FALSE, main = title)
}

par(mfrow = c(1, 2))
plot_digit(xtest[i_fp, ], title = "False Positive: predicted 1, true 7")
par(mfrow = c(1, 1))

```


# Compare training vs test accuracy

```{r}

pred_1nn_train <- knn(train = xtrain, test = xtrain, cl = ytrain, k = 1)
pred_27nn_train <- knn(train = xtrain, test = xtrain, cl = ytrain, k = 27)

acc_1nn_train <- mean(pred_1nn_train == ytrain)
acc_27nn_train <- mean(pred_27nn_train == ytrain)
acc_1nn_test <- conf_1nn$overall['Accuracy']
acc_27nn_test <- conf_27nn$overall['Accuracy']

cat("\n1NN - Training Accuracy:", round(acc_1nn_train, 4), 
    "| Test Accuracy:", round(acc_1nn_test, 4), "\n")
cat("27NN - Training Accuracy:", round(acc_27nn_train, 4), 
    "| Test Accuracy:", round(acc_27nn_test, 4), "\n\n")

```

# We end this study by summarize as the followinng 
```{r}
cat("\nSUMMARY\n")
summary_df <- data.frame(
  Model = c("1NN", "27NN"),
  Accuracy = c(round(acc_1nn_test, 4), round(acc_27nn_test, 4)),
  AUC = c(round(auc_1nn, 4), round(auc_27nn, 4)),
  Sensitivity = c(round(conf_1nn$byClass['Sensitivity'], 4), 
                  round(conf_27nn$byClass['Sensitivity'], 4)),
  Specificity = c(round(conf_1nn$byClass['Specificity'], 4), 
                  round(conf_27nn$byClass['Specificity'], 4))
)
print(summary_df)
```









#Exercise 3 : Discovering the concept of ultra high dimen-sionality

```{r}
# install.packages("mlbench")
library(mlbench)
library(corrplot)
library(ggplot2)
library(gridExtra)
library(car)
options(warn = -1)
```

```{r}


path1 = "/home/guest/Desktop/JSD/SML/Dataset/"

data("DNA")          
data("BreastCancer") 

#different data set
spam      <- read.csv(paste0(path1, "spam-classification-1.csv"), header = TRUE)
leukemia  <- read.csv(paste0(path1, "leukemia-data-1.csv"),       header = TRUE)
prostate  <- read.csv(paste0(path1, "prostate-cancer-1.csv"),     header = TRUE)
colon     <- read.csv(paste0(path1,"colon-cancer-1.csv"),        header = TRUE)


```

#let's define the main function that can take any data set and do all necessaries analysis
```{r}
options(warn = -1)

analyze_dataset <- function(data, name, target_col = NULL, exclude_cols = NULL) {
  cat("\n====================", name, "====================\n\n")
  
  if (!is.null(exclude_cols)) data <- data[, !(names(data) %in% exclude_cols)]
  if (!is.null(target_col) && target_col %in% names(data)) {
    target <- data[[target_col]]
    data   <- data[, names(data) != target_col]
    cat("Target:", target_col, "\n\n")
  }
  
  
  data_clean <- na.omit(data)
  n <- nrow(data_clean); p <- ncol(data_clean); kappa <- n / p
  
  cat("1) Dimension\n")
cat("   n =", n, ", p =", p, ", $\\kappa = n/p =", round(kappa, 4), "$\n")
  if (kappa > 5)      cat("   → Classical: plenty of samples per variable\n\n")
  else if (kappa>=1) cat("   → Moderate: regularization helpful\n\n")
  else               cat("   → Ultra high-dimensional (p > n): strong regularization / feature selection needed\n\n")
  
  cat("2) Type and scale homogeneity\n")
  types <- sapply(data_clean, function(x) class(x)[1])
  print(table(types))
  num_numeric <- sum(sapply(data_clean, is.numeric))
  cat("   Numeric vars:", num_numeric, "\n")
  if (length(unique(types)) == 1) cat("   → All features same type\n") else cat("   → Mixed feature types\n")
  
  
  num_cols <- sapply(data_clean, is.numeric)
  if (any(num_cols)) { 
    num_data <- data_clean[, num_cols, drop = FALSE]
    ranges <- sapply(num_data, function(x) diff(range(x, na.rm = TRUE)))
    range_ratio <- max(ranges) / (min(ranges) + 1e-10)
    cat("   Range ratio (max/min):", round(range_ratio, 2), "\n")
    if (range_ratio < 10) cat("   → Scales broadly comparable\n\n") else cat("   → Scales very different (standardize)\n\n")
  } else {
    cat("   No numeric variables\n\n")
  }
  
  cat("3) Graphical summaries (up to 9 vars)\n\n")
  p_plot <- min(9, p)
  set.seed(42)
  sel <- if (p > 9) sample(1:p, 9) else 1:p
  plots <- vector("list", p_plot)
  for (i in seq_len(p_plot)) {
    vname <- names(data_clean)[sel[i]]
    vdat  <- data_clean[[sel[i]]]
    df    <- data.frame(x = vdat)
    if (is.numeric(vdat)) {
      plots[[i]] <- ggplot(df, aes(x = x)) +
        geom_histogram(bins = 30, fill = "steelblue", color = "black") +
        labs(title = vname, x = "", y = "Freq") +
        theme_minimal(base_size = 9)
    } else {
      plots[[i]] <- ggplot(df, aes(x = x)) +
        geom_bar(fill = "coral", color = "black") +
        labs(title = vname, x = "", y = "Count") +
        theme_minimal(base_size = 9) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
    }
  }
  grid.arrange(grobs = plots, ncol = 3, top = paste(name, "- distributions"))
  
  cat("4) Correlations and multicollinearity\n\n")
  num_cols <- sapply(data_clean, is.numeric)
  if (sum(num_cols) > 1) {
    num_data <- data_clean[, num_cols, drop = FALSE]
    if (ncol(num_data) > 50) {
      cat("   Using first 50 numeric variables for correlation analysis\n")
      num_data <- num_data[, 1:50]
    }
    C <- cor(num_data, use = "pairwise.complete.obs")
    high <- which(abs(C) > 0.7 & abs(C) < 1, arr.ind = TRUE)
    cat("   Pairs with |r| > 0.7:", nrow(high) / 2, "\n")
    if (nrow(high) > 0) {
      cat("   → Strong multicollinearity present\n")
      upper <- high[high[,1] < high[,2], , drop = FALSE]
      vals  <- C[upper]
      top   <- order(abs(vals), decreasing = TRUE)[1:min(5, length(vals))]
      cat("   Top correlated pairs:\n")
      for (k in top) {
        i <- upper[k,1]; j <- upper[k,2]
        cat("     ", colnames(C)[i], "<->", colnames(C)[j], 
            ": r =", round(C[i,j], 3), "\n")
      }
    } else cat("   → No strong multicollinearity\n")
    cat("\n")
    
    if (ncol(num_data) <= 20) {
      corrplot(C, method = "color", type = "upper",
               tl.cex = 0.7, tl.col = "black",
               title = paste(name, "- correlation matrix"),
               mar = c(0,0,2,0))
    } else {
      idx <- 1:20
      corrplot(C[idx, idx], method = "color", type = "upper",
               tl.cex = 0.7, tl.col = "black",
               title = paste(name, "- correlation (first 20 vars)"),
               mar = c(0,0,2,0))
    }
  } else {
    cat("   Not enough numeric variables for correlations\n\n")
  }
}


```

#Apply on all the dataset
```{r}

data(DNA)
analyze_dataset(DNA, "DNA", target_col = "Class")

data(BreastCancer)
analyze_dataset(BreastCancer, "BreastCancer",
                target_col = "Class", exclude_cols = "Id")

cat("\nSpam Classification\n")
tryCatch({
  spam <- read.csv(paste0(path1,"spam-classification-1.csv"))
  analyze_dataset(spam, "Spam", target_col = names(spam)[ncol(spam)])
}, error = function(e) cat("  File spam-classification-1.csv not found.\n"))

cat("\nLeukemia\n")
tryCatch({
  leukemia <- read.csv(paste0(path1,"leukemia-data-1.csv"))
  analyze_dataset(leukemia, "Leukemia", target_col = names(leukemia)[1])
}, error = function(e) cat("  File leukemia-data-1.csv not found.\n"))

cat("\nProstate cancer\n")
tryCatch({
  prostate <- read.csv(paste0(path1,"prostate-cancer-1.csv"))
  analyze_dataset(prostate, "Prostate", target_col = names(prostate)[1])
}, error = function(e) cat("  File prostate-cancer-1.csv not found.\n"))

cat("\nColon cancer\n")
tryCatch({
  colon <- read.csv(paste0(path1,"colon-cancer-1.csv"))
  analyze_dataset(colon, "Colon", target_col = names(colon)[1])
}, error = function(e) cat("  File colon-cancer-1.csv not found.\n"))
```


#Exercise BOnus 
#1)
#2)


```{r}
set.seed(123)
n <- 99

X <- runif(n, 0, 2*pi)

Y <- numeric(n)
for(i in 1:n) {
  mean_y <- (pi/2) * X[i] + (3*pi/4) * cos((pi/2) * (1 + X[i]))
  sd_y <- pi / sqrt(18)
  Y[i] <- rnorm(1, mean = mean_y, sd = sd_y)
}

cat("Toatl samples of (X_i, Y_i) : ", n, "\n\n")
```

#3 scatterplot

```{r}
plot(X, Y, 
     main = "Scatterplot of Generated Data",
     xlab = "X", 
     ylab = "Y",
     pch = 16, 
     col = rgb(0, 0, 1, 0.5),
     cex = 0.8)


x_seq <- seq(0, 2*pi, length.out = 200)
f_star <- (pi/2) * x_seq + (3*pi/4) * cos((pi/2) * (1 + x_seq))
lines(x_seq, f_star, col = "red", lwd = 2)
legend("topright", legend = c("Data", "True f*(X)"), 
       col = c("blue", "red"), pch = c(16, NA), lty = c(NA, 1), lwd = c(NA, 2))
```
# Question 4a: Find f*(X)
#Question 4b: Find R* (Bayes Risk)
# Question 4c: Compare different regression methods

```{r}
train_size <- round(0.6 * n)
train_idx <- sample(1:n, train_size)
test_idx <- setdiff(1:n, train_idx)

X_train <- X[train_idx]
Y_train <- Y[train_idx]
X_test <- X[test_idx]
Y_test <- Y[test_idx]

# we will replied n_rep
n_rep <- 100
test_errors <- matrix(NA, nrow = n_rep, ncol = 4)
colnames(test_errors) <- c("kNN", "Linear", "Polynomial", "Tree")
```


```{r}

# libraries
if(!require(FNN)) install.packages("FNN", repos = "http://cran.rstudio.com/")
if(!require(rpart)) install.packages("rpart", repos = "http://cran.rstudio.com/")
library(FNN)
library(rpart)
```


```{r}
for(rep in 1:n_rep) {
  # Generate new sample
  X_new <- runif(n, 0, 2*pi)
  Y_new <- numeric(n)
  for(i in 1:n) {
    mean_y <- (pi/2) * X_new[i] + (3*pi/4) * cos((pi/2) * (1 + X_new[i]))
    sd_y <- pi / sqrt(18)
    Y_new[i] <- rnorm(1, mean = mean_y, sd = sd_y)
  }
  
  # Split
  X_train_new <- X_new[train_idx]
  Y_train_new <- Y_new[train_idx]
  X_test_new <- X_new[test_idx]
  Y_test_new <- Y_new[test_idx]
  
  # 1. kNN Regression (k=5)
  pred_knn <- knn.reg(train = matrix(X_train_new), 
                      test = matrix(X_test_new), 
                      y = Y_train_new, 
                      k = 5)$pred
  test_errors[rep, 1] <- mean((Y_test_new - pred_knn)^2)
  # 2. Linear Regression
  model_linear <- lm(Y_train_new ~ X_train_new)
  pred_linear <- predict(model_linear, newdata = data.frame(X_train_new = X_test_new))
  test_errors[rep, 2] <- mean((Y_test_new - pred_linear)^2)
  
  # 3. Polynomial Regression (degree 5)
  model_poly <- lm(Y_train_new ~ poly(X_train_new, degree = 5, raw = TRUE))
  pred_poly <- predict(model_poly, newdata = data.frame(X_train_new = X_test_new))
  test_errors[rep, 3] <- mean((Y_test_new - pred_poly)^2)
  
  # 4. Regression Tree
  df_train <- data.frame(X = X_train_new, Y = Y_train_new)
  df_test <- data.frame(X = X_test_new)
  model_tree <- rpart(Y ~ X, data = df_train, control = rpart.control(cp = 0.01))
  pred_tree <- predict(model_tree, newdata = df_test)
  test_errors[rep, 4] <- mean((Y_test_new - pred_tree)^2)
}
```




# comparative boxplots
```{r}
par(mar = c(5, 4, 4, 2) + 0.1)
boxplot(test_errors, 
        main = "Test Errors Comparison (100 Replications)",
        ylab = "Mean Squared Error",
        xlab = "Method",
        col = c("lightblue", "lightgreen", "lightcoral", "lightyellow"),
        las = 1)
abline(h = pi^2/18, col = "red", lwd = 2, lty = 2)
legend("topright", legend = "Bayes Risk R*", col = "red", lty = 2, lwd = 2)
```

# Question 4d: Comment on results

```{r}
cat("\nQuestion 4d: Commentary\n")
cat("Average Test Errors:\n")
print(colMeans(test_errors))
cat("\nBayes Risk R* =", pi^2/18, "\n\n")

```




